dispRity metrics

### 2017-01-25 doc/dispRity-ecology-demo.Rmd

This demo aims to give quick overview of the `dispRity` package (v.0.2) for ecological analysis.
Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more details.

To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, one might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.

@@

For this example with ecological data we are going to use data from McClean (unpubl.).
This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.

@@

## As well as two list of different factors affecting each experimental plot:
## the treatment and the depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
```

@@

```{r, fig.width=6, fig.height=6}
## The x and y axis represent the two first dimensions (i.e. principal
## components) of the ordination
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

@@

Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.

@@

The first step in such analysis will be to create different series that are two sub-samples of the ordinated space (i.e. the *n*-dimensional object).
Each of these series contain a certain number of elements (i.e. the 40 experimental field plots) that have some attributes in common.
In our example, we are going to group the elements according to their depth and treatment.

@@

```{r}
## Splitting the ordinated space into four subsamples
customised_series <- cust.series(ord_matrix, factors)

## Note that the output of dispRity functions are dispRity objects
class(customised_series)
## These objects are automatically printed in a summary method (calling S3 print.dispRity)
## giving information about the object
customised_series
```

For more details on the `dispRity` objects, see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.pdf).
Basically the idea is to avoid jamming the `R` console such as when using:

```{r}
## Summarising the object
str(customised_series)

## Displaying the full object
print(customised_series, all = TRUE)
```

@@

Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity *as defined by the user*.
For more details on measuring disparity, see the [`dispRity` metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.Rmd).

In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and
centroid.

```{r}
## Calculating disparity
disparity <- dispRity(customised_series, metric = c(median, centroids))

## Note that disparity is a dispRity object and printing it just gives details
## on the object, not the results. We need to use summary.dispRity (S3) to get
## the results.
disparity
summary(disparity)
```

@@

One might also want to bootstrap the data to test the robustness of the measured disparity to outliers.
Also, as we can see, each series has different numbers of elements.
It might be interesting to rarefy the data as well to have only series with the same number of elements.
Both steps are easily doable through the `boot.matrix` function.

```{r}
## Simple bootstrapping (100 times)
bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)

## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
rarefied_data <- boot.matrix(customised_series, bootstraps = 100,
    rarefaction = 17)

## Note that the output is a dispRity object giving some details on the series
## and the bootstraps
bootstrapped_data
rarefied_data
```

@@

```{r}
## Bootstrapped disparity
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))

## Rarefied disparity
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
```

@@

We can now summarise the data using various options such as the confidence intervals levels and the central tendency.

```{r}
## Summarising disparity (default)
summary(disparity_BS)

## The quantiles are calculated as 50 and 95 and the central tendency is the
## mean by default but we can specify different options
summary(disparity_BS, quantile = 90, cent.tend = median)

## Finally we can see the results of the rarefaction analysis:
summary(disparity_rare)

## Note that for the first three series, there is no observed disparity since
## the data was rarefied for these series.
```

We can also plot the results and have a look at the effect of the number of experimental plots:

```{r, fig.width=6, fig.height=6}
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")

>>>>

This demo aims to give quick overview of the `dispRity` package (v.0.3) for ecological analysis.
Please refer to [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes, namely the `dispRity` manual that explains the functions in more details.

To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, we might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.

@@

For this example with ecological data we are going to use data from D. McClean (unpubl.).
This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.

@@

## As well as two list of different factors affecting each experimental plot: treatment and depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
```

@@

```{r, fig.width=6, fig.height=6}
## The x and y axis represent the two first dimensions (principal components) of the ordination
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## Setting the colors that will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## Setting the symbols that will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

@@

Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.

@@

The first step in such analysis will be to create different series that will represent sub-samples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
Each of these series contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
In our example, we are going to group the elements according to their depth and treatment.

@@

```{r}
## Splitting the ordinated space into four subsamples
(customised_series <- cust.series(ord_matrix, factors))
```

Note that the output of `dispRity` functions are `dispRity` objects.
They allow specific attributes to be displayed using the basic S3 functions (`print`, `summary`, `plot`, etc...).
For more details on the `dispRity` objects, see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html).

@@

Disparity can be calculated in many ways, and therefore, the `dispRity` function allows to measure disparity as defined by the user.
For more details on measuring disparity, see the [`dispRity` metric vignette](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metrics.html).

In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the `dispRity` function by feeding them to the `metric` argument.
Here we are going to feed the functions `stats::median` and `dispRity::centroids` which calculates distances between elements and their centroid.

```{r}
## Calculating disparity as the median distance between each elements and their group centroid
(disparity <- dispRity(customised_series, metric = c(median, centroids)))
```

Similarly than with `cust.series`, `dispRity` displays a `dispRity` object.
To actually look at the calculated values, we can summarise the data by simply using `summary`:

```{r}
## Displaying the summary of the calculated disparity
summary(disparity)
```

@@

We can also want to bootstrap the data to test the robustness of the measured disparity to outliers.
Also, as we can see, each series has different numbers of elements.
It might be interesting to rarefy the data as well to have only series with the same number of elements.
Both steps are easily doable through the `boot.matrix` function.

```{r}
## Bootstrapping the customised series 100 times
(bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100))

## Bootstrapping the series and using rarefaction (i.e. only re-sampling 17 elements each time)
(rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17))
```

@@

```{r}
## Calculating the bootstrapped disparity
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))

## Calculating the rarefied disparity
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
```

@@

We can now summarise the data using various options such as the confidence intervals levels and the central tendency.
For that we will use the `summary` function mentioned above using several of the options specific to `dispRity` objects (see more details with `?summary.dispRity`).

```{r}
## Summarising bootstrapped disparity
summary(disparity_BS)

## Summarising the rarefied disparity
summary(disparity_rare)

## Summarising the same results but using the mean as the central tendency and displaying 
## the 90th quantile only.
summary(disparity_rare, quantile = 90, cent.tend = mean)
```

We can also plot the results and have a look at the effect of the number of experimental plots using the generic `plot` function and it's associated arguments (see more details with `?plot.dispRity` for the arguments specific to `dispRity` objects).

```{r, fig.width=12, fig.height=6}
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")

### 2017-01-25 doc/dispRity-metrics.Rmd

Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.Rmd)), the `metric` argument has to follow certain rules:

>>>>

Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
However, because of the infrastructure of the package (especially the one of the `dispRity` objects – see the [`dispRity` manual](http://htmlpreview.github.com/?https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-manual.html), the `metric` argument has to follow certain rules:
